---
title: "Stat545 Homework 1"
author: "Eric Sanders"
date: 'Submitted for 2018-09-18'
header-includes:
   - \usepackage{bbm}
   - \usepackage{amsmath}
   - \usepackage{amsfonts}
   - \usepackage{amssymb}
   - \usepackage{caption}
   - \usepackage{tikz}
   - \usepackage{textcomp}
   - \usepackage{float}
   - \usepackage{setspace}
   - \doublespacing
   - \allowdisplaybreaks
   - \pagenumbering{roman}
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

In this document we present some data exploration completed using the `ChickWeight` dataset.

# Loading in and Examining Data Numerically

First, we load the data

```{r}
# Call to preloaded data frame
data(ChickWeight)
```

We will look at three variables in this data set. Namely, we will look at

* `Diet`, which is categorical and marks four different chick diets.
* `Weight`, which is continuous and indicates how large the chicks are growing.
* `Age`, which is numerical and marks the 12 different ages at which chicks have measured.

Let us see how this data set looks.

```{r}
# Keep only required columns
ChickWeight = ChickWeight[,c('Diet','Time','weight')]
# Check structure, see first six rows
str(ChickWeight)
head(ChickWeight)
# Identify number of observations
nrow(ChickWeight)
```

So we can see that there are `r nrow(ChickWeight)` observations of these three variables.

# Exploring the Data with Plots

Now, let us look at how weight tends to vary with age. We can even include some jitter to get a better idea of how the points are dispersed.

```{r}
plot(data=ChickWeight, weight~jitter(Time,1), pch=16,
     cex=0.6, xlim=c(0,22), ylim=c(40,375), xlab="Age",
     ylab="Weight", main="Weight vs. Age")
```

Unfortunately, this doesn't help very much if we are interested in comparing the diets of the chicks. Let us try to make four plots for the four diets with the same axes in order to compare.

We can also add to each plot a lowess fitted line. This is not for the sake of any inference or prediction, but just to consolidate our visualization into some loosely fitted lines.

```{r}
# First declare four subsets of the data set -- this will make future code less bulky when we split into one of the for diets for a plot.
ChickWeight1 = ChickWeight[which(ChickWeight$Diet==1), ]
ChickWeight2 = ChickWeight[which(ChickWeight$Diet==2), ]
ChickWeight3 = ChickWeight[which(ChickWeight$Diet==3), ]
ChickWeight4 = ChickWeight[which(ChickWeight$Diet==4), ]

# Prepare to plot 4 plots next to each other
par(mfrow=c(2,2))

# Create plot for Diet 1
plot(data=ChickWeight1,
     weight~jitter(Time, 1), col='blue', pch=16, cex=0.6, 
     xlim=c(0, 22), ylim=c(40, 375), xlab="Age", ylab="Weight",
     main="Diet 1,  Weight vs. Age")
# Add lowess line to Diet 1 plot
lines(lowess(ChickWeight1$Time,
             ChickWeight1$weight))
# Create plot for Diet 2
plot(data=ChickWeight2,
     weight~jitter(Time, 1), pch=16, cex=0.6, col="red",
     xlim=c(0, 22), ylim=c(40, 375), xlab="Age", ylab="Weight",
     main="Diet 2,  Weight vs. Age")
# Add lowess line to Diet 2 plot
lines(lowess(ChickWeight2$Time,
             ChickWeight2$weight))
# Create plot for Diet 3
plot(data=ChickWeight3,
     weight~jitter(Time, 1), pch=16, cex=0.6, col="green",
     xlim=c(0, 22), ylim=c(40, 375), xlab="Age", ylab="Weight",
     main="Diet 3,  Weight vs. Age")
# Add lowess line to Diet 3 plot
lines(lowess(ChickWeight3$Time,
             ChickWeight3$weight))
# Create plot for Diet 4
plot(data=ChickWeight4,
     weight~jitter(Time, 1), pch=16, cex=0.6, col="purple",
     xlim=c(0, 22), ylim=c(40, 375), xlab="Age", ylab="Weight",
     main="Diet 4,  Weight vs. Age")
# Add lowess line to Diet 4 plot
lines(lowess(ChickWeight4$Time,
             ChickWeight4$weight))
```

As an alternative method of comparing diets, we can add these four fitted lines to the original plot, as follows.

```{r}
# Create plot with only blue points (Diet 1)
plot(data=ChickWeight1,
     weight~jitter(Time, 1), col='blue', pch=16, cex=0.6,
     xlim=c(0, 22), ylim=c(40, 375), xlab="Age", 
     ylab="Weight", main="Weight vs. Age by Diet")
# Add red points (Diet 2)
points(ChickWeight2$weight~
         jitter(ChickWeight2$Time),
       pch=16, cex=0.6, col="red")
# Add green points (Diet 3)
points(ChickWeight3$weight~
         jitter(ChickWeight3$Time),
       pch=16, cex=0.6, col="green")
# Add purple points (Diet 4)
points(ChickWeight4$weight~
         jitter(ChickWeight4$Time),
       pch=16, cex=0.6, col="purple")
# Add all four lowess lines
lines(lowess(ChickWeight1$Time,
             ChickWeight1$weight),
      col='blue', lwd=2)
lines(lowess(ChickWeight2$Time,
             ChickWeight2$weight),
      col='red', lwd=2)
lines(lowess(ChickWeight3$Time,
             ChickWeight3$weight),
      col='green', lwd=2)
lines(lowess(ChickWeight4$Time,
             ChickWeight4$weight),
      col='purple', lwd=2)
# Add legend labelling colours
legend(x=0, y=350, col=c('blue', 'red', 'green', 'purple'),
       lty=c(1, 1, 1, 1), c("Diet 1", "Diet 2", "Diet 3", 'Diet 4'))
```

After exploring the data, we may wish to further investigate if there is anything tangible to the observation that Diet 3 seems to have a higher contribution to weight in older chicks.